---
title: "Analysis of Literature Coding"
author: "Mandy Norrbo"
date: "29/01/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Required packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```

### Loading in the data
```{r, message = FALSE}
dat <- read_csv("quantdata.csv")
```

### Prevalence of internal meta-analyses
```{r}

#calculating how many articles used internal meta-analysis out of all coded empirical articles
nmini <- dat %>% 
  filter(EMPIRICAL == "yes") %>% 
  count(USES_MINIMETA == "yes") %>%
  arrange(n)
  
```

`r nmini$n[1]` out of `r nmini$n[2]+nmini$n[1]` empirical research articles published between May and September 2019 in the Journal of Experimental Social Psychology used an internal meta-analysis — that is `r round(nmini$n[1]/(nmini$n[2]+nmini$n[1])*100)`%.

### Cleaning up data
```{r}

tidydat <- dat %>%
  #selecting columns of interest in analysis
  select(-c("TITLE", "DOI", "EMPIRICAL", "SIGTEST", "COMMENT")) %>% 
  #selecting the studies that should be included in the control vs minimeta analysis
  filter(INCLUDE == "yes") %>% 
  select(-"PVAL8", -"PVAL9", -"PVAL10", -"PVAL11", -"PVAL12", -"N8", -"N9", -"N10", -"N11", -"N12") %>% #filtering out empty columns 
  gather(key = "PVALNO", value = "PVALUES", PVAL1,PVAL2,PVAL3,PVAL4,PVAL5,PVAL6, PVAL7)

```

### How many multi-study papers?
```{r}
#no of articles with more than 1 study
multidat <- dat %>% 
  select(STUDY_NO) %>% 
  filter(STUDY_NO > 1) %>% 
  nrow()

#no of articles with just 1 study
singledat <- dat %>% 
  select(STUDY_NO) %>% 
  filter(STUDY_NO == 1) %>% 
  nrow()
```

`r multidat` out of `r singledat+multidat` empirical articles reported the findings of more than one study, only `r singledat` reported the findings from a single study.

### Number of studies combined in meta-analysis
```{r}

#selecting articles that used minimeta
minidat <- tidydat %>%  
  filter(USES_MINIMETA == "yes")

#number of individual studies that were aggregated
nstudymini <- minidat  %>%
  summarise(mean = mean(META_STUDY_NO),
            median = median(META_STUDY_NO),
            sd = sd(META_STUDY_NO))

```

The number of studies combined in the mini meta-analyses coded in this overview range from `r min(minidat$META_STUDY_NO)` to `r max(minidat$META_STUDY_NO)`, with a mean of `r nstudymini$mean`, median of `r nstudymini$median`, and with a standard deviation of `r round(nstudymini$sd, 2)`.

### Plotting the p-values
```{r, warning = FALSE}

#labels for plot
supp.labs <- c("Used internal meta-analysis", "Did not use internal meta-analysis")

g1 <- ggplot(tidydat, aes(x=PVALNO, y=PVALUES)) +
  geom_point(aes(colour = STUDY_ID)) +
  geom_line(aes(group = STUDY_ID, colour = STUDY_ID)) + 
  geom_segment(aes(x = 0, xend = 7, y =.05, yend = 0.05), colour = "red", size = 0.5, linetype = "longdash") +
  facet_wrap(~CONTROL, labeller = labeller(CONTROL = supp.labs)) +
  theme(axis.text.x  = element_text(size = 5)) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_discrete(labels=c("PVAL1" = "1", "PVAL2" = "2", 
                            "PVAL3" = "3", "PVAL4" = "4", 
                            "PVAL5"= "5", "PVAL6" = "6", "PVAL7" = "7"), name = "Study") +
  scale_y_continuous(name = "p-value") +
  labs(title = "p-value distribution across studies by article type") +
  theme(plot.title = element_text(size = 15, face = "bold"), strip.text.x = element_text(size = 17), axis.title = element_text(size = 15))

ggsave("pvalueplot.png", width = 10, height = 6)

```

### How many p > 0.05?

```{r}

#how many p-values > 0.05 grouped by minimeta use (yes/no)
nullp <- tidydat %>%
  filter(PVALUES > 0.05) %>% 
  group_by(USES_MINIMETA) %>% 
  summarise(n = n())

#how many individual studies in total in 20 articles
studytot <- tidydat %>% 
  filter(PVALUES != "") %>% 
  nrow()


```

Out of the 20 articles and `r studytot` individual studies, `r nullp$n[1]+nullp$n[2]` p-values were nonsignificant. Out of these, `r nullp$n[2]` were in the articles that used internal meta-analyses, while `r nullp$n[1]` were in the matched control articles that did not end up using an internal meta-analysis.


### Sample sizes

```{r}

#sample size dataframe for plotting & descriptives
ndat <- dat %>% 
  filter(INCLUDE == "yes", USES_MINIMETA == "yes") %>% 
  select(c("N1", "N2", "N3", "N4", "N5", "N6", "N7", "USES_MINIMETA", "STUDY_ID")) %>% 
  gather(key = "NSTUDY", value = "NVALUES", 
         N1, N2, N3, N4, N5, N6, N7)

#difference between largest and smallest sample sizes within an article across individual studies
bystudy <- ndat %>% 
  group_by(STUDY_ID) %>% 
  summarise(min = min(NVALUES, na.rm = TRUE), max = max(NVALUES, na.rm = TRUE)) %>% 
  mutate(difference = max - min)

```

Sample sizes within single studies ranged from `r min(ndat$NVALUES, na.rm = TRUE)` to `r max(ndat$NVALUES, na.rm = TRUE)`, with a mean size of `r round(mean(ndat$NVALUES, na.rm = TRUE))` and median size of `r median(ndat$NVALUES, na.rm = TRUE)`.

### Plotting sample sizes

```{r, warning = FALSE}

g2 <- ggplot(ndat, aes(x=NSTUDY, y=NVALUES)) +
  geom_point(aes(colour = STUDY_ID)) +
  geom_line(aes(group = STUDY_ID, colour = STUDY_ID)) + 
  theme(axis.text.x  = element_text(size = 5)) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_discrete(labels=c("N1" = "1", "N2" = "2", 
                            "N3" = "3", "N4" = "4", 
                            "N5"= "5", "N6" = "6", "N7" = "7"), name = "Study") +
  scale_y_continuous(name = "Total sample size (N)") +
  labs(title = "Sample size distribution across studies using internal meta-analysis") +
  theme(plot.title = element_text(size = 15, face = "bold"), axis.title = element_text(size = 15))


ggsave("nplot.png", width = 10, height = 6)
```

### Pre-registrations

```{r}

#dataframe with prereg information
prereg <- dat %>% filter(INCLUDE == "yes", USES_MINIMETA == "yes") %>% select(STUDY_ID, STUDY_NO,PRE_REG) 

#percentage pre-registered
percreg <- round(sum(prereg$PRE_REG)/sum(prereg$STUDY_NO),2)*100
#total no. pre-registered
totalreg <- sum(prereg$PRE_REG)
#out of how many studies
outof <- sum(prereg$STUDY_NO)

```


Within the internal meta-analysis articles, `r totalreg` out of `r outof`, , that is `r percreg`%, were reported to have been pre-registered. Only 1 article (82_2) seemed to have pre-registered all studies within the internal meta-analysis. However, the internal meta-analysis itself did not seem to be pre-registered and thus, when to run it and which studies to include was still flexible. 

### How many do not report p-values?

```{r}

#articles without p-values were coded as "not sure" in "IS_MINI_SIG" variable
nopval <- dat %>% 
  filter(IS_MINI_SIG == "not sure") %>% 
  nrow()

```


Out of the `r nmini$n[1]` coded articles using an internal meta-analysis, `r nopval` did not report an overall significance level (i.e. p-value) for the meta-analysis, instead reporting only effect sizes and confidence intervals.
